Investigating Red Wine dataset by Adam McCarthy
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(ggthemes)
library(scales)
library(extRemes)
library(sm)
library(gridExtra)
library(GGally)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
setwd('./data')
# Load the Data
red_wine <- read.csv('wineQualityReds.csv')
setwd('..')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Process_the_Data}
# Create a new categorical variable for quality
red_wine['quality_c'] <- ifelse(red_wine$quality <= 5, "lower", "upper")

# To be able to represent quality as a categorical variable
#
# Assign a full range of levels to represent 0-10
# red_wine$quality <- factor(red_wine$quality)
# levels(red_wine$quality) <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10)

# Assign categorical groups as levels
red_wine$quality_c <- factor(red_wine$quality_c)
```

> **Dataset**: Red wine dataset consiting of 1599 variants of the Portuguese "Vinho Verde" wine. Includes physicochemical (inputs) and sensory (the output) variables. This dataset is public available for research. The details are described in [Cortez et al., 2009].

# Univariate Plots Section

> **Overview**: There are 13 variables within the original dataset. Of these one is a
dependent variable giving a subjective measure of quality based on experts
sensory reviw of the wine. The main 11 variables are independent physiochemical
tests. These may be inter-related but are initially thought of as individual
measurements.

See [here](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt) for more information.

  
 * X - Each row of data, each wine type tested.    - Index
 
 Input variables (based on physicochemical tests):
 
 * 1 - fixed acidity, tartaric acid - $(g/dm^{3})$
 * 2 - volatile acidity, acetic acid - $(g/dm^{3})$
 * 3 - citric acid - $(g/dm^{3})$
 * 4 - residual sugar - $(g/dm^{3})$
 * 5 - chlorides, sodium chloride - $(g/dm^{3})$
 * 6 - free sulfur dioxide $(mg/dm^{3})$
 * 7 - total sulfur dioxide $(mg/dm^{3})$
 * 8 - density $(g/cm^{3})$
 * 9 - pH
 * 10 - sulphates, potassium sulphate - $(g/dm^{3})$
 * 11 - alcohol (% by volume)

 Output variable (based on sensory data):
 
 * 12 - quality (score between 0 and 10)
 * 13 - quality_c (quality grouped into lower and upper)

```{r echo=FALSE, Univariate_Plots}
summary(red_wine)
```

The summary table depicts descriptive statistics for all of the variables.

Quality is measured from 0-10 based. To plot effectively each 
element from 0-10 is included to ensure the representation of data 
is better presented. 
This variable has been reformatted to a factor with 11 levels (from orignal 6).

One categorical variables within the dataset is generated to
group quality into two categories. Lower (<=5) and Upper (>5).
This will be used to explore for any major differences in values.

Ph is on a scale of 0 to 14 (potential of hydrogen), 
all values here are acidic (<7) with the range
occuring between 2.7-4.

Alcholol is measured in percentages.

The other variables are all numeric data types.

There are no missing values within the dataset.

```{r echo=FALSE, Univariate_Plots_Quality, fig.height = 3, fig.width = 5, fig.align="center"}
ggplot(data=red_wine, aes(red_wine$quality)) + 
  geom_bar(fill="#9099A2") +
  theme_tufte() +
  labs(
    title="Distribution of Quality"
    ) +
  labs(
    x="Quality",
    y="Count"
    )
```

Quality is between 2 and 8 despite the scoring range being 0-10. The majority
of values are 5 and 6. There is skew to the right tail giving more positive
scores in 7. Giving the poor sampling of scores outside of 5 and 6 the quality
measure will be split into two categories, upper (>5) and lower (<=5).

```{r echo=FALSE, Univariate_Plots_multi_2, fig.height = 3, fig.width = 5, fig.align="center"}
small_hist <- function(data_v, name, scale, df=red_wine) {
plt <- ggplot(data = df, aes(data_v)) + 
  geom_histogram(bins = 50,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = name
    ) +
  labs(
    x = scale,
    y = "Count"
    )

return(plt)
}

plt10 <- small_hist(red_wine$alcohol, 
  "Alcohol", "%")

plt10
```

Alcohol also has a skewed distribution of values but with most values located
between 9% and 13%. A small histogram can give a quick impression of the
distribution of data more effectively than reading the descriptive statistics.

```{r echo=FALSE, Multiplot_function}
# Entirely taken from:
# http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

```{r echo=FALSE, Univariate_Plots_multi, fig.height = 6, fig.width = 9}
plt1 <- small_hist(red_wine$fixed.acidity, "Fixed Acidity", "g/dm^3")
plt2 <- small_hist(red_wine$volatile.acidity, "Volatile Acidity", "g/dm^3")
plt3 <- small_hist(red_wine$citric.acid, "Citric Acidity", "g/dm^3")
plt4 <- small_hist(red_wine$residual.sugar, "Residual Sugar", "g/dm^3")
plt5 <- small_hist(red_wine$chlorides, "Chlorides", "g/dm^3")
plt6 <- small_hist(red_wine$pH, "pH", "pH scale")
plt7 <- small_hist(red_wine$free.sulfur.dioxide, 
  "Free Sulphar Dioxide", "mg/dm^3")
plt8 <- small_hist(red_wine$total.sulfur.dioxide, 
  "Total Sulphar Dioxide", "mg/dm^3")
plt9 <- small_hist(red_wine$density, 
  "Density", "g/dm^3")

multiplot(plt1, plt2, plt3, plt4, plt5, plt6, plt7, plt8, plt9, cols = 3)
```

Starting to investigate further variables a group of 9 measurements are placed
into histograms to understand their distributions. Of these pH, Fixed acidity
and Volatile Acidity have slightly normal distributions, these could be 
investigated further using quantile quantile plots.

Residual Sugar and Chlorides both appear to have highly skewed, over dispersed
data, these could be explored using a transformation to investigate the
distribution of values further.

Citric Acidity has two spikes, one occuring at 0 and one around 0.5. Bin size
could be changed to investigate if there is some artifact in the data related
to rounding.

Density appears relatively normal but 
with occasional spikes, these should be checked for rounding issues.

Free sulphar dioxide and total sulphar dioxide have skewed distributions, with
a sparse population of values towards the maximum.

```{r echo=FALSE, hist_pH, fig.height = 4, fig.width = 6, fig.align="center"}
hist(red_wine$pH,
     prob=TRUE,
     breaks=60,
     col='#9099A2',
     border="white",
     col.axis = '#9099A2',
     col.lab = '#9099A2',
     col.main = '#192231',
     bty = 'n')
lines(density(red_wine$pH),
      col='#192231', lwd=2)
lines(density(red_wine$pH, adjust=2), 
      lty="dotted", col='#192231', lwd=1)
rug(red_wine$pH, ticksize=0.02, side=1, lwd=0.2)
```

Two plots detailing pH value.

```{r echo=FALSE, QQ_pH, fig.height = 8, fig.width = 8, fig.align="center"}

qqnorm(red_wine$pH,
              main = 'pH QQ plot',
              bty = 'n',  # Frame style
              col = alpha('#192231', 0.3),  # transparency with col
              cex = .4,  # scale pof points
              col.axis = '#9099A2',
              col.lab = '#9099A2',
              col.main = '#192231',
              pch = 20  # point symbol type
)
qqline(red_wine$pH, col = '#9099A2', lty = 1)
```

Now using these two plots pH can be examined in greater detail. pH appears to 
have light tails as displayed by the overall sigmoidal shape of the points, 
it displays very fine grained clustering which is related to rounding of 
values and it shows an overall skew to the right.

Towards the upper right of the quantile quantile plot against a standard 
normal distribution the line crosses the 95% confidence intervals.

```{r echo=FALSE, QQ_den, fig.height = 4, fig.width = 4, fig.align="center"}

qqnorm(red_wine$density,
              main = 'Density QQ plot',
              bty = 'n',  # Frame style
              col = alpha('#192231', 0.3),  # transparency with col
              cex = .4,  # scale pof points
              col.axis = '#9099A2',
              col.lab = '#9099A2',
              col.main = '#192231',
              pch = 20  # point symbol type
)
qqline(red_wine$density, col = '#9099A2', lty = 1)
```

Density also shows light tails and skew to the right leading to the fit for
a normal distribution exisiting outside of the 95% confidence intervals.

```{R echo=FALSE, Log_hist, fig.height = 4, fig.width = 5}
plt11 <- ggplot(aes(x = red_wine$residual.sugar), data =red_wine) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "Residual sugar"
    ) +
  labs(
    x = "g/dm^3",
    y = "Count"
    )
plt12 <- plt11 + scale_x_sqrt() +
    labs(title = "Residual sugar sqrt", x = "g/dm^3 sqrt")
plt13 <- plt11 + scale_x_log10() +
    labs(title = "Residual sugar log10", x = "g/dm^3 log10")

grid.arrange(plt11, plt12, plt13, ncol = 1)
```

To explore over disperssed variables data transformations can be applied.
The original plot followed by a square root and log10 data transformation is conducted on residual sugar in the above three plots. The data transformations
give a better representation of the distribution of values. They still have
long tails towards the right.

```{R echo=FALSE, Log_hists, fig.height = 6, fig.width = 5}
plt14 <- ggplot(aes(x = red_wine$chlorides), data = red_wine) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "Chlorides log10"
    ) +
  labs(
    x = "g/dm^3 log10",
    y = "Count"
    ) + scale_x_log10()

plt15 <- ggplot(aes(x = red_wine$citric.acid), data = red_wine) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "Citric Acid sqrt"
    ) +
  labs(
    x = "g/dm^3 sqrt",
    y = "Count"
    ) + scale_x_sqrt()

plt16 <- ggplot(aes(x = red_wine$total.sulfur.dioxide), data = red_wine) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "Total sulfur dioxide log10"
    ) +
  labs(
    x = "g/dm^3 log10",
    y = "Count"
    ) + scale_x_log10()

plt17 <- ggplot(aes(x = red_wine$free.sulfur.dioxide), data = red_wine) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "Free Sulfur Dioxide sqrt"
    ) +
  labs(
    x = "g/dm^3 sqrt",
    y = "Count"
    ) + scale_x_sqrt()

grid.arrange(plt14, plt15, plt16, plt17, ncol = 1)
```

Other over disperssed variables are explored in the above plot and each is
matched to a data transformation that better represents 
the distribution of data values. Chlorides shows a slightly normal distribution
with a long tail to the right, it is centered nicely after a 
log10 transformation.

Citric acid still has a large count of values at close to 0, the data
distribution is best represented using a square root transformation, this
does not bring the data closer to a normal distribution but represents
the the distribution of values across the range.

Total sulfur dioxide is transformed using log10 given a wide normal
styled distribution.

Free sulfure dioxide puts too many values to the right hand side during a
log10 transformation so even with a skew, square root gives a better
distribution of values.

Each of these transformations can be used when comparing these variables
to other variables.

# Univariate Analysis

### What is the structure of your dataset?
The dataset is in a tidy format. Each observation corresponds to a series
of variables.

The dataset consists of one key dependent variable, quality. This is based on
the subjective sensory assesment of experts. This should result in a value
between 0-10, within this dataset the majority of the samples exist with
results of 5 or 6.

The other eleven variables are all measurements. It is assumed that these
are indpendent, although there may be relationships between some of these
evelen variables.

### What is/are the main feature(s) of interest in your dataset?
The main feauture of interest is the dependent variable quality.

Without further exploratory work it can not be assesed which of the
11 measurement variables is the most important.

### What other features in the dataset do you think will help support your \
At the moment all other features are of interest as this investigation is
not applying any domain knowledge about what is most important.

### Did you create any new variables from existing variables in the dataset?
One categorical variable is created based on the dependent quality variable.
This splits quality into two categories, upper and lower. The purpose of this
is to represent better (upper) and worse (lower) wines. This is split in two
due to the limited distribution of values outside of 5 and 6. The idea is to
use this new variable to see if there are any relationships in other variables
that seperate better or worse wines.

### Of the features you investigated, were there any unusual distributions?
There appear to be no normal distributions in this dataset, two are close
but the qq plots have shown they sit outside of an idealised normal
distribution.

Rounding of inputs appears to cause minor clustering within the datasets,
this may relate to the tool precision used to assess the physiochemical
properties of each wine.

Citric acid has a large number of values close to zero, this should be
further investigated to see if this is a data quality issue or true signal.

Variables can depict over disperssed distributions with long tails to the right
, using either square root or log10 transformations can help better
represent the distribtion of these values. Each of these as been investigated
to identify which transformation should be applied for future plots.


# Bivariate Plots Section

This section will begin with assesing if there any variables with a strong
difference between upper and lower wine quality reviews.

This section will also check if any of the measurement variables are related.

```{r echo=FALSE, Bivariate_Plots, fig.height = 2, fig.width = 4, fig.align="center"}
freq_poly <-  function(x_v, name, scale, 
                       dataset=red_wine, group=red_wine$quality_c) {
plt <- qplot(x = x_v, data = dataset,
             geom = 'freqpoly', 
             color = group,
             bins = 25) +
  scale_x_continuous() +
  theme_tufte() +
  labs(
    title = name
    ) +
  labs(
    x = scale
      )
plt + scale_color_manual(values=c('#0B3C5D', '#D9B310'))
}

freq_poly(red_wine$alcohol, "Alcohol frequency polygon", "g/dm^3")
```

Frequency polygons showing the distribution of values but now split
by groups of the dependent variable quality can help show if there
are trends in any of the variables and the output variables.

In the above figure alcohol shows a large count of wines with lower
quality measurements are associated to lower alcohol values.

```{r echo=FALSE, freq_poly_mulit, fig.height = 6, fig.width = 9, message=FALSE, warning=FALSE,}
# Use function to create a range of plots
plt19 <- freq_poly(red_wine$fixed.acidity, "Fixed Acidity", "g/dm^3")
plt20 <- freq_poly(red_wine$volatile.acidity, "Volatile Acidity", "g/dm^3")
plt21 <- freq_poly(red_wine$citric.acid, "Citric Acidity", "g/dm^3 sqrt")
plt21 <- plt21 + scale_x_sqrt()

plt22 <- freq_poly(red_wine$residual.sugar, "Residual Sugar", "g/dm^3 log10")
plt22 <- plt22 + scale_x_log10()

plt23 <- freq_poly(red_wine$chlorides, "Chlorides", "g/dm^3 log10")
plt23 <- plt23 + scale_x_log10()

plt24 <- freq_poly(red_wine$pH, "pH", "pH scale")
plt25 <- freq_poly(red_wine$free.sulfur.dioxide, 
  "Free Sulphar Dioxide", "mg/dm^3 sqrt")
plt25 <- plt25 + scale_x_sqrt()

plt26 <- freq_poly(red_wine$total.sulfur.dioxide, 
  "Total Sulphar Dioxide", "mg/dm^3 log10")
plt26 <- plt26 + scale_x_log10()

plt27 <- freq_poly(red_wine$density, 
  "Density", "g/dm^3")

multiplot(plt19, plt20, plt21, plt22, plt23, 
          plt24, plt25, plt26, plt27, cols = 3)
```

By plotting the remaining variables in a similar way it is possible to
quickly identify any variables that appear to have some relationship
to the upper or lower quality. The first observation is that many of the
variables show little difference between the frequency polygons for upper or
for lower. Those that do show a some difference, but it does not appear to
be large or very obvious.

Volatile acidity shows a distribution centered to the left for upper. The
frequency polygon shows lower values for upper than the lower category.

Total sulfar dioxide has none of it´s upper values in the upper group.

Both free sulfar dioxide and total sulfar dioxide show some two peaks for
the upper group (this may relate to count as this is not a density plot).

Citric acid has a spike in it´s higher values for the upper group.

These issues will be worth investigating further.

```{r echo=FALSE, freq_pair_plot, fig.height = 9, fig.width = 9, message=FALSE, warning=FALSE,}
ggpairs(red_wine,
        title = "Red Wine Pair Plot"
        ) +
  theme_minimal()
```

The pair plot gives all combinations of bivariate analysis. It´s main
limitation is that it is not created using the data transformations
previously identified. This acts as a usefull way to look up
pairs of values to investigate further.

As previously addressed by the bivariate frequency polygon plot very
few of the datasets impact the upper and lower categories of quality
significantly apart from alcohol. This can be viewed in the box plots and
paired histograms. Alcohol has a 0.48 correlation to quality on this plot.
Volatile acidity has a negative 0.39 and sulphates has 0.25. These are
variables with the highest correlation to the quality variable.

Free sulphar and total sulphar have a correlation of 0.67 but the two
variables have little correlation to other variables.

Density appears to have some weak correlation with a few variables like
residual sugar and fixed acidity. This could be a candidate for
multi-variate analysis.

Chlorides and residual sugars should be checked as these are both highly
dispersed variables so it is difficult to see any correlation in this pair
plot.

```{r echo=FALSE, scatter_sulf, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$total.sulfur.dioxide, y = red_wine$free.sulfur.dioxide),
       data = red_wine) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  coord_trans(x = 'log10', y = 'log10') +
  theme_tufte() +
  labs(
    title = "Total vs Residual Sulfar Dioxide",
    x = "Total mg/dm^3 log10",
    y = "Residual mg/dm^3 log10"
    ) +
  geom_smooth(formula = log(y)~log(x),
              colour = '#192231')
```

Total vs. residual sulfar dioxide is plotted with both variables on a log10
scale to highlight the correlation between the two variables. There are a
handfull of outliers but this appears to show a relationship.

```{r echo=FALSE, scatter_den, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$density, y = red_wine$fixed.acidity),
       data = red_wine) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Density vs Fixed Acidity",
    x = "Density g/dm^3",
    y = "Fixed acidity mg/dm^3"
    ) +
  geom_smooth(method = 'lm', formula = y~x,
              colour = '#192231')
```

Density plotted against fixed acidity shows a relationship between the two
variables. This has been fitted with a linear model.

```{r echo=FALSE, scatter_c_r, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$chlorides, y = red_wine$residual.sugar),
       data = red_wine) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  coord_trans(x = 'log10', y = 'log10') +
  theme_tufte() +
  labs(
    title = "Chlorides vs Residual Sugar",
    x = "Chlorides g/dm^3 log10",
    y = "Residual sugar g/dm^3 log10"
    )
```

Chlorides and residual sugar are compared with a log10 transform but there
still appears to limited correlation between these two variables.

```{r echo=FALSE, scatter_acid, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$volatile.acidity, y = red_wine$fixed.acidity),
       data = red_wine) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Volatile vs Fixed Acidity",
    x = "Volatile acidity g/dm^3",
    y = "Fixed acidity mg/dm^3"
    ) +
  geom_smooth(method = 'lm', formula = y~x,
              colour = '#192231')
```

The fixed compared to volatile acidity shows a weak correlation.

```{r echo=FALSE, scatter_q_v, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$sulphates, y = red_wine$quality),
       data = red_wine) +
  geom_point(alpha = 0.3,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Sulphates vs Quality",
    x = "Sulphates g/dm^3",
    y = "Quality score"
    )
```

Comparing sulphates to quality values there intitially appears to be a
correlated. In the following series of plots category 5 and 6 are always key
as they have a much higher proportion of values. Category 5 and 6 have a wider
distribution of values of than the other categories. Despite the apparent
trend between high quality and low quality values this may not be signal given
the low number of samples in the highest and lowest values. Still there is a
complete difference between categories 3 sulphate values and category 8
sulphate values.

Linear models are not used in these plots as they will always be swayed by
the high number of values within quality values 5 and 6.

```{r echo=FALSE, scatter_q_v, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$volatile.acidity, y = red_wine$quality),
       data = red_wine) +
  geom_point(alpha = 0.3,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Volatility Acidity vs Quality",
    x = "Volatile acidity g/dm^3",
    y = "Quality score"
    )
```

Comparing the dependent variable, quality, to variables
that have a correlation to it starts with volatile acidity.
The plot shows each category of quality on the y axis and then a
scatter of plots showing the distribution of volatile acidity. There is a weak
correlation but what is more interesting is which values do not occur in
certain classes. For example there are no values above 0.9 for the top
scored wines. For both the high and low quality scores the challenge is
in the low sample of variables.

```{r echo=FALSE, scatter_q_a, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$alcohol, y = red_wine$quality),
       data = red_wine) +
  geom_point(alpha = 0.3,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Alcohol vs Quality",
    x = "Alcohol %",
    y = "Quality score"
    )
```

Alcohol compared to quality shows a positive correlation. The values
are dominatly within the 6 and 5 scores. Lowest scoring wines often have
lower alcohol values. While the highest scores tend towards the higher values.
A key observation is cateogry six which shows a range of values across the
alcohol values. This puts into question the strength of this correlation
that higher alcohol equals higher quality of wine.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Overall there are few strong correlations between variables in this dataset.
The most promising line of investigation seems to be alcohol and it´s
relationship to quality.

Two other variables (Volatile acidity and sulphates) have a weak relationship
to the feature of interest.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Few strong relationships exist, density appears to have associations to
multiple variables and can be investigated further through multivariate
analysis.

A number of plots have outliers, it would be interesting to know if there
are clusters within the dataset or any other types of structures that can
not be observed through bivariate analysis.

### What was the strongest relationship you found?

Total and free sulfar dioxide have the strongest relationship found,
this would make sense as it is likely free sulfar dioxide has
a proportional relationship to the total amount of sulfar
dioxide in a sample.

Density to fixed acidity has the same correlation value in the pair plot
as total and free sulfar dioxide. 

# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}
ggplot(aes(x = red_wine$density, y = red_wine$fixed.acidity),
       data = red_wine) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = red_wine$alcohol) +
  theme_tufte() +
  labs(
    title = "Densiy vs Fixed Acidity with Alcohol",
    x = "Density g/dm^3 log10",
    y = "Fixed Acidity g/dm^3"
    )
```

The

```{r echo=FALSE, Multivariate_Plots_2}
ggplot(aes(x = red_wine$alcohol, y = red_wine$quality_c),
       data = red_wine) +
  geom_point(alpha = 0.8,
             position = position_jitter(),
             color = red_wine$quality) +
  theme_tufte() +
  labs(
    title = "Densiy vs Residual Sugar with Alcohol",
    x = "Alcohol %",
    y = "Residual sugar g/dm^3 log10"
    )
```

The plot

```{r echo=FALSE, Multivariate_Plots_3}
ggplot(aes(x = red_wine$alcohol, y = red_wine$volatile.acidity),
       data = red_wine) +
  geom_point(alpha = 0.8,
             position = position_jitter(),
             color = red_wine$quality) +
  theme_tufte() +
  labs(
    title = "Densiy vs Residual Sugar with Alcohol",
    x = "Alcohol %",
    y = "Residual sugar g/dm^3 log10"
    )
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!

## Dataset from:
P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
Modeling wine preferences by data mining from physicochemical properties.
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.
Available at: 
[@Elsevier](http://dx.doi.org/10.1016/j.dss.2009.05.016)
[Pre-press (pdf)](http://www3.dsi.uminho.pt/pcortez/winequality09.pdf)
[bib](http://www3.dsi.uminho.pt/pcortez/dss09.bib)