Exploratory Data Analysis of Red Wine
========================================================

### *Adam McCarthy*

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(ggthemes)
library(scales)
library(extRemes)
library(sm)
library(gridExtra)
library(GGally)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
setwd('..')
setwd('./data')
# Load the Data
red_wine <- read.csv('wineQualityReds.csv')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Process_the_Data}
# Create a new categorical variable for quality
red_wine['upper_vs_lower'] <- ifelse(red_wine$quality <= 5, "lower", "upper")

# Assign categorical groups as levels
red_wine$upper_vs_lower <- factor(red_wine$upper_vs_lower)

# Remove X as this is the same as index
red_wine$X <- NULL
```

Red wine data-set consisting of 1599 variants of the Portuguese 
"Vinho Verde" wine. Includes physicochemical (inputs) and sensory 
(the output) variables. This data-set is public available for research. 
The details are described in [Cortez et al., 2009].

# Univariate Plots Section

> **Overview**: There are 13 variables within the original data-set. 
Of these one is a
dependent variable giving a subjective measure of quality based on experts
sensory review of the wine. The main 11 variables are independent 
physio-chemical
tests. These may be inter-related but are initially thought of as individual
measurements.
  
 * X - Each row of data, each wine type tested.
 
 Input variables (based on physio-chemical tests):
 
 * 1 - fixed acidity, tartaric acid - $(g/dm^{3})$
 * 2 - volatile acidity, acetic acid - $(g/dm^{3})$
 * 3 - citric acid - $(g/dm^{3})$
 * 4 - residual sugar - $(g/dm^{3})$
 * 5 - chlorides, sodium chloride - $(g/dm^{3})$
 * 6 - free sulfur dioxide $(mg/dm^{3})$
 * 7 - total sulfur dioxide $(mg/dm^{3})$
 * 8 - density $(g/cm^{3})$
 * 9 - pH
 * 10 - sulphates, potassium sulphate - $(g/dm^{3})$
 * 11 - alcohol (% by volume)

 Output variable (based on sensory data):
 
 * 12 - quality (score between 0 and 10)
 * 13 - upper_vs_lower (quality grouped into lower and upper)

              **Descriptive Statistics of Variables**
```{r echo=FALSE, Univariate_Plots}
summary(red_wine)
```

The summary table depicts descriptive statistics for all of the variables.

Quality is measured from 0-10 based. However the data-set is limited to
values between 3 and 8.

One categorical variables within the data-set is generated to
group quality into two categories. Lower (<=5) and Upper (>5).
This will be used to explore for any major differences in values.

pH is on a scale of 0 to 14 (potential of hydrogen), 
all values here are acidic (<7) with the range
occurring between 2.7-4.

Alcohol is measured in percentages.

The other variables are all numeric data types.

There are no missing values within the data-set.

```{r echo=FALSE, U, fig.height = 3, fig.width = 9, fig.align="center"}
ggplot(data = red_wine, aes(red_wine$quality)) + 
  geom_bar(fill = "#9099A2") +
  theme_tufte() +
  labs(
    title = "Distribution of Quality"
    ) +
  labs(
    x = "Quality",
    y = "Count"
    ) +
  scale_x_continuous(breaks = round(seq(min(red_wine$quality),
                                        max(red_wine$quality), by = 1),1))
```

Quality is between 2 and 8 despite the scoring range being 0-10. The majority
of values are 5 and 6. There is skew to the right tail giving more positive
scores in 7. Giving the poor sampling of scores outside of 5 and 6 the quality
measure will be split into two categories, upper (>5) and lower (<=5).

```{r echo=FALSE, U2, fig.height = 3, fig.width = 5, fig.align="center"}
small_hist <- function(data_v, name, scale, df=red_wine) {
plt <- ggplot(data = df, aes(data_v)) + 
  geom_histogram(bins = 50,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = name
    ) +
  labs(
    x = scale,
    y = "Count"
    )

return(plt)
}

plt10 <- small_hist(red_wine$alcohol, 
  "Alcohol", "%")

plt10
```

Alcohol also has a skewed distribution of values but with most values located
between 9% and 13%. A small histogram can give a quick impression of the
distribution of data more effectively than reading the descriptive statistics.

```{r echo=FALSE, Multiplot_function}
# Entirely taken from:
# http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots == 1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

```{r echo=FALSE, Univariate_Plots_multi, fig.height = 6, fig.width = 9}
plt1 <- small_hist(red_wine$fixed.acidity, "Fixed Acidity", "g/dm^3")
plt2 <- small_hist(red_wine$volatile.acidity, "Volatile Acidity", "g/dm^3")
plt3 <- small_hist(red_wine$citric.acid, "Citric Acidity", "g/dm^3")
plt4 <- small_hist(red_wine$residual.sugar, "Residual Sugar", "g/dm^3")
plt5 <- small_hist(red_wine$chlorides, "Chlorides", "g/dm^3")
plt6 <- small_hist(red_wine$pH, "pH", "pH scale")
plt7 <- small_hist(red_wine$free.sulfur.dioxide, 
  "Free Sulphar Dioxide", "mg/dm^3")
plt8 <- small_hist(red_wine$total.sulfur.dioxide, 
  "Total Sulphar Dioxide", "mg/dm^3")
plt9 <- small_hist(red_wine$density, 
  "Density", "g/dm^3")

# Multiplot used when grid.arrange does not work (e.g. not a ggplot)
multiplot(plt1, plt2, plt3, plt4, plt5, plt6, plt7, plt8, plt9, cols = 3)
```

Investigate further, 9 variables are placed
into histograms to understand their distributions. Of these pH, Fixed acidity
and Volatile Acidity have slightly normal distributions, these could be 
investigated further using quantile quantile plots.

Residual Sugar and Chlorides both appear to have highly skewed, over dispersed
data, these could be explored using a transformation to investigate the
distribution of values further.

Citric Acidity has two spikes, one occurring at 0 and one around 0.5. Bin size
could be changed to investigate if there is some artifact in the data related
to rounding.

Density appears relatively normal but 
with occasional spikes, these should be checked for rounding issues.

Free sulfur dioxide and total sulfur dioxide have skewed distributions, with
a sparse population of values towards the maximum.

```{r echo=FALSE, hist_pH, fig.height = 4, fig.width = 6, fig.align="center"}
hist(red_wine$sulphates,
     prob = TRUE,
     breaks = 60,
     col = '#9099A2',
     border = "white",
     col.axis = '#9099A2',
     col.lab = '#9099A2',
     col.main = '#192231',
     bty = 'n',
     xlab = "Sulphates g/dm^3",
     ylab = "Density"
     )
lines(density(red_wine$sulphates),
      col = '#192231', lwd = 2)
lines(density(red_wine$sulphates, adjust = 2), 
      lty = "dotted", col = '#192231', lwd = 1)
rug(red_wine$sulphates, ticksize = 0.02, side = 1, lwd = 0.2)
```

Sulphates also has over dispersed values to the right with a 
skewed distribution.
The rug plot helps highlight the low occurrence of these values in two clusters,
close to 2 and close to 1.6.

```{r echo=FALSE, QQ_pH, fig.height = 8, fig.width = 8, fig.align="center"}

qqnorm(red_wine$pH,
              main = 'pH QQ plot',
              bty = 'n',  # Frame style
              col = alpha('#192231', 0.3),  # transparency with col
              cex = .4,  # scale pof points
              col.axis = '#9099A2',
              col.lab = '#9099A2',
              col.main = '#192231',
              pch = 20  # point symbol type
)
qqline(red_wine$pH, col = '#9099A2', lty = 1)
```

Using this plot, pH can be examined in greater detail and
compared to an idealized normal distribution. pH appears to 
have light tails as displayed by the overall sigmoidal shape of the points, 
it displays very fine grained clustering which is related to rounding of 
values and it shows an overall skew to the right.

Towards the upper right of the quantile quantile plot against a standard 
normal distribution the line crosses the 95% confidence intervals.

```{r echo=FALSE, QQ_den, fig.height = 8, fig.width = 8, fig.align="center"}

qqnorm(red_wine$density,
              main = 'Density QQ plot',
              bty = 'n',  # Frame style
              col = alpha('#192231', 0.3),  # transparency with col
              cex = .4,  # scale pof points
              col.axis = '#9099A2',
              col.lab = '#9099A2',
              col.main = '#192231',
              pch = 20  # point symbol type
)
qqline(red_wine$density, col = '#9099A2', lty = 1)
```

Density also shows light tails and skew to the right. The fit for
a normal distribution is outside of the 95% confidence intervals.

```{R echo=FALSE, Log_hist, fig.height = 4, fig.width = 5}
plt11 <- ggplot(aes(x = red_wine$residual.sugar), data = red_wine) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "Residual sugar"
    ) +
  labs(
    x = "g/dm^3",
    y = "Count"
    )
plt12 <- plt11 + scale_x_sqrt() +
    labs(title = "Residual sugar sqrt", x = "g/dm^3 sqrt")
plt13 <- plt11 + scale_x_log10() +
    labs(title = "Residual sugar log10", x = "g/dm^3 log10")

grid.arrange(plt11, plt12, plt13, ncol = 1)
```

To improve visualization of over dispersed variables 
data transformations can be applied.
The original plot followed by a square root and log10 
data transformation is conducted on residual sugar in the above three plots. 
The data transformations
give a better representation of the distribution of values. They still have
long tails towards the right in this case.

```{R echo=FALSE, Log_hists, fig.height = 6, fig.width = 5}
plt14 <- ggplot(aes(x = red_wine$chlorides), data = red_wine) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "Chlorides log10"
    ) +
  labs(
    x = "g/dm^3 log10",
    y = "Count"
    ) + scale_x_log10()

plt15 <- ggplot(aes(x = red_wine$citric.acid), data = red_wine) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "Citric Acid sqrt"
    ) +
  labs(
    x = "g/dm^3 sqrt",
    y = "Count"
    ) + scale_x_sqrt()

plt16 <- ggplot(aes(x = red_wine$total.sulfur.dioxide), data = red_wine) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "Total sulfur dioxide log10"
    ) +
  labs(
    x = "g/dm^3 log10",
    y = "Count"
    ) + scale_x_log10()

plt17 <- ggplot(aes(x = red_wine$free.sulfur.dioxide), data = red_wine) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "Free Sulfur Dioxide sqrt"
    ) +
  labs(
    x = "g/dm^3 sqrt",
    y = "Count"
    ) + scale_x_sqrt()

plt18 <- ggplot(aes(x = red_wine$sulphates), data = red_wine) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "Sulphates log10"
    ) +
  labs(
    x = "g/dm^3 log10",
    y = "Count"
    ) + scale_x_log10()

grid.arrange(plt14, plt15, plt16, plt17, plt18, ncol = 1)
```

Other over dispersed variables are explored in the above plot and each is
matched to a data transformation that better represents 
the distribution of data values. Chlorides shows a slightly normal distribution
with a long tail to the right, it is centered nicely after a 
log10 transformation.

Citric acid still has a large count of values at close to 0, the data
distribution is best represented using a square root transformation, this
does not bring the data closer to a normal distribution but represents
the the distribution of values across the range.

Total sulfur dioxide is transformed using log10 given a wide normal
styled distribution.

Free sulfur dioxide puts too many values to the right hand side during a
log10 transformation so even with a skew, square root gives a better
distribution of values.

Sulphates gives a better distribution of values but is still skewed.

Each of these transformations can be considered
when comparing these variables
to other variables in further stages of the analysis.

# Univariate Analysis

### What is the structure of your dataset?
The data-set is in a tidy format. Each observation corresponds to a series
of variables.

The data-set consists of one key dependent variable, quality. This is based on
the subjective sensory assessment of experts. This should result in a value
between 0-10, within this data-set the majority of the samples exist with
results of 5 or 6.

The other eleven variables are all measurements. It is assumed that these
are independent, although there may be relationships between some of these
eleven variables.

### What is/are the main feature(s) of interest in your dataset?
The main feature of interest is the dependent variable quality.

Without further exploratory work it can not be assessed which of the
11 measurement variables is the most important.

### What other features in the dataset do you think will help support your \
investigation?

At the moment all other features are of interest as this investigation is
not applying any domain knowledge about what is most important.

### Did you create any new variables from existing variables in the dataset?
One categorical variable is created based on the dependent quality variable.
This splits quality into two categories, upper and lower. The purpose of this
is to represent better (upper) and worse (lower) wines. This is split in two
due to the limited distribution of values outside of 5 and 6. The idea is to
use this new variable to see if there are any relationships in other variables
that separate better or worse wines.

### Of the features you investigated, were there any unusual distributions?
There appear to be no normal distributions in this data-set, two are close
but the qq plots have shown they sit outside of an idealized normal
distribution.

Rounding of inputs appears to cause minor clustering within the data-sets,
this may relate to the tool precision used to assess the physio-chemical
properties of each wine.

Citric acid has a large number of values close to zero, this could be
further investigated to see if this is a data quality issue or true signal.

Variables can depict over dispersed distributions with long tails to the right
, using either square root or log10 transformations can help better
represent the distribution of these values. Each of these as been investigated
to identify which transformation should be applied for future plots.


# Bivariate Plots Section

This section will begin with assessing if there any variables with a strong
difference between upper and lower wine quality reviews.

This section will also check if any of the measurement variables are related.

```{r echo=FALSE, BPlots, fig.height = 2, fig.width = 4, fig.align="center"}
freq_poly <-  function(x_v, name, scale, 
                       dataset=red_wine, group=red_wine$upper_vs_lower) {
plt <- qplot(x = x_v, data = dataset,
             geom = 'freqpoly', 
             color = group,
             bins = 25) +
  scale_x_continuous() +
  theme_tufte() +
  labs(
    title = name
    ) +
  labs(
    x = scale
      )
plt + scale_color_manual(values = c('#0B3C5D', '#D9B310'))
}

freq_poly(red_wine$alcohol, "Alcohol frequency polygon", "g/dm^3")
```

Frequency polygons showing the distribution of values but now split
by groups of the dependent variable quality can help show if there
are trends in any of the variables and the output variables.

In the above figure alcohol shows a large count of wines with lower
quality measurements are associated to lower alcohol values.

```{r echo=F, B2, fig.height = 2, fig.width = 4, fig.align="center", message=F, warning=F}
sul <- freq_poly(red_wine$sulphates, "Sulphates frequency polygon",
                 "g/dm^3 log10")
sul <- sul + scale_x_log10()
sul
```

Sulphates also displays a difference between the upper and lower categories.
In which higher values of sulphates appear to exclude the lower quality
scores.

```{r echo=FALSE, fm, fig.height = 6, fig.width = 9, message=F, warning=F,}
# Use function to create a range of plots
plt19 <- freq_poly(red_wine$fixed.acidity, "Fixed Acidity", "g/dm^3")
plt20 <- freq_poly(red_wine$volatile.acidity, "Volatile Acidity", "g/dm^3")
plt21 <- freq_poly(red_wine$citric.acid, "Citric Acidity", "g/dm^3 sqrt")
plt21 <- plt21 + scale_x_sqrt()

plt22 <- freq_poly(red_wine$residual.sugar, "Residual Sugar", "g/dm^3 log10")
plt22 <- plt22 + scale_x_log10()

plt23 <- freq_poly(red_wine$chlorides, "Chlorides", "g/dm^3 log10")
plt23 <- plt23 + scale_x_log10()

plt24 <- freq_poly(red_wine$pH, "pH", "pH scale")
plt25 <- freq_poly(red_wine$free.sulfur.dioxide, 
  "Free Sulfurr Dioxide", "mg/dm^3 sqrt")
plt25 <- plt25 + scale_x_sqrt()

plt26 <- freq_poly(red_wine$total.sulfur.dioxide, 
  "Total Sulfur Dioxide", "mg/dm^3 log10")
plt26 <- plt26 + scale_x_log10()

plt27 <- freq_poly(red_wine$density, 
  "Density", "g/dm^3")

multiplot(plt19, plt20, plt21, plt22, plt23, 
          plt24, plt25, plt26, plt27, cols = 3)
```

By plotting the remaining variables in a similar way it is possible to
quickly identify any variables that appear to have a stronger relationship
to the upper or lower quality group. The first observation is that many of the
variables show little difference between the frequency polygons for upper or
for lower. Those that do show some difference, but it does not appear to
be large or very obvious at first inspection.

Volatile acidity shows a distribution centered to the left for upper. The
frequency polygon shows lower values for upper than the lower category.

Total sulfur dioxide has none of it´s upper values in the upper group.

Both free sulfur dioxide and total sulfur dioxide show some two peaks for
the upper group (this may relate to count as this is not a density plot).

Citric acid has a spike in it´s higher values for the upper group.

These issues will be worth investigating further.

```{r echo=FALSE, freq_pair_plot, fig.height = 9, fig.width = 9, message=FALSE, warning=FALSE,}
ggpairs(red_wine,
        title = "Red Wine Pair Plot"
        ) +
  theme_minimal()
```

The pair plot gives all combinations of bi-variate analysis. It´s main
limitation is that it is not created using the data transformations
previously identified. This acts as a useful way to look up
pairs of values to investigate further.

As previously addressed by the bi-variate frequency polygon plot very
few of the data-sets impact the upper and lower categories of quality
significantly apart from alcohol. This can be viewed in the box plots and
paired histograms. Alcohol has a 0.48 correlation to quality on this plot.
Volatile acidity has a negative 0.39 and sulphates has 0.25. These are
variables with the highest correlation to the quality variable.

Free sulfur and total sulfur have a correlation of 0.67 but the two
variables have little correlation to other variables.

Density appears to have some weak correlation with a few variables like
residual sugar and fixed acidity. This could be a candidate for
multi-variate analysis.

Chlorides and residual sugars should be checked as these are both highly
dispersed variables so it is difficult to see any correlation in this pair
plot.

An alternative way to view which variables are correlated is with a correlation
matrix. This represents similar observations as above but the magnitude
of the colour highlights strong correlations.

```{r echo=FALSE, correlation_matrix, fig.height = 5, fig.width = 5, message=FALSE, warning=FALSE}
ggcorr(red_wine)  +
  theme_tufte()
```

```{r echo=FALSE, scatter_sulf, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$total.sulfur.dioxide, y = red_wine$free.sulfur.dioxide),
       data = red_wine) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  coord_trans(x = 'log10', y = 'log10') +
  theme_tufte() +
  labs(
    title = "Total vs Residual Sulfur Dioxide",
    x = "Total mg/dm^3 log10",
    y = "Residual mg/dm^3 log10"
    ) +
  geom_smooth(formula = log(y)~log(x),
              colour = '#192231')
```

Total vs. residual sulfur dioxide is plotted with both variables on a log10
scale to highlight the correlation (0.67) between the two variables. 
There are a handful of outliers but this appears to show a relationship.

```{r echo=FALSE, scatter_den, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$density, y = red_wine$fixed.acidity),
       data = red_wine) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Density vs Fixed Acidity",
    x = "Density g/dm^3",
    y = "Fixed acidity mg/dm^3"
    ) +
  geom_smooth(method = 'lm', formula = y~x,
              colour = '#192231')
```

Density plotted against fixed acidity shows a relationship between the two
variables. This has been fitted with a linear model.

```{r echo=FALSE, scatter_den2, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$density, y = red_wine$alcohol),
       data = red_wine) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Density vs Alcohol",
    x = "Density g/dm^3",
    y = "Alcohol %"
    )
```

Density compared to alcohol shows a negative relationship, it does not
fit well using a linear model. There is a cluster of values with low
alcohol. It shows a good spread of values.

```{r echo=FALSE, scatter_c_r, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$chlorides, y = red_wine$residual.sugar),
       data = red_wine) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  coord_trans(x = 'log10', y = 'log10') +
  theme_tufte() +
  labs(
    title = "Chlorides vs Residual Sugar",
    x = "Chlorides g/dm^3 log10",
    y = "Residual sugar g/dm^3 log10"
    )
```

Chlorides and residual sugar are compared with a log10 transform but there
still appears to be limited correlation between these two variables.

```{r echo=FALSE, scatter_sulp, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$sulphates, y = red_wine$volatile.acidity),
       data = red_wine) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Sulphates vs Volatile Acidity",
    x = "Sulphates g/dm^3",
    y = "Volatile Acidity g/dm^3"
    ) +
  coord_trans(x = 'log10', y = 'log10')
```

The comparison of sulphates to volatile acidity shows a weak negative
relationship. It shows a good dispersion of values when both axis
are on log10. A linear model does not fit well to this plot.

```{r echo=FALSE, scatter_acid, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$volatile.acidity, y = red_wine$citric.acid),
       data = red_wine) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Volatile Acidity vs Citric Acid",
    x = "Volatile acidity g/dm^3",
    y = "Citric Acid g/dm^3"
    ) +
  geom_smooth(method = 'lm', formula = y~x,
              colour = '#192231')
```

The fixed compared to volatile acidity shows a negative correlation.

```{r echo=FALSE, scatter_acid2, fig.height = 5, fig.width = 5}
ggplot(aes(x = red_wine$citric.acid, y = red_wine$fixed.acidity),
       data = red_wine) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Citric Acid vs Fixed Acidity",
    x = "Citric acid g/dm^3",
    y = "Fixed acidity g/dm^3"
    ) +
  geom_smooth(method = 'lm', formula = y~x,
              colour = '#192231')
```

Citric vs fixed acidity shows a positive relationship.

```{r echo=FALSE, scatter_q_v, fig.height = 5, fig.width = 5}
ggplot(aes(y = red_wine$sulphates, x = factor(red_wine$quality)),
       data = red_wine) +
  geom_boxplot(outlier.colour = "transparent",
               color = "#9099A2") +
  coord_flip() +
  geom_jitter(alpha = 0.3,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Sulphates vs Quality",
    y = "Sulphates g/dm^3",
    x = "Quality score"
    ) + 
  scale_y_continuous(breaks = round(seq(min(red_wine$quality),
                                        max(red_wine$quality), by = 1),1))
```

Comparing sulphates to quality values gives an initial appearance of a
correlation. In the following series of plots category 5 and 6 are always key
as they have a much higher proportion of values. Category 5 and 6 have a wider
distribution of values of than the other categories. Despite the apparent
trend between high quality and low quality values this may not be signal given
the low number of samples in the highest and lowest values. Still there is a
complete difference between categories 3 sulphate values and category 8
sulphate values.

Linear models are not used in these plots as they will always be swayed by
the high number of values within quality values 5 and 6.

```{r echo=FALSE, scatter_q_v2, fig.height = 5, fig.width = 5}
ggplot(aes(y = red_wine$volatile.acidity, x = factor(red_wine$quality)),
       data = red_wine) +
  geom_boxplot(outlier.colour = "transparent",
               color = "#9099A2") +
  coord_flip() +
  geom_jitter(alpha = 0.3,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Volatility Acidity vs Quality",
    y = "Volatile acidity g/dm^3",
    x = "Quality score"
    )  + 
  scale_y_continuous(breaks = round(seq(min(red_wine$quality),
                                        max(red_wine$quality), by = 1),1))
```

Comparing the dependent variable, quality, to variables
that have a correlation to it includes volatile acidity.
The plot shows each category of quality on the y axis and then a
scatter of plots showing the distribution of volatile acidity. There is a weak
correlation but what is more interesting is which values do not occur in
certain classes. For example there are no values above 0.9 for the top
scored wines. For both the high and low quality scores the challenge is
in the low sample of variables.

```{r echo=FALSE, scatter_q_a, fig.height = 5, fig.width = 5}
finplt1 <-  ggplot(aes(y = red_wine$alcohol, x = factor(red_wine$quality)),
                       data = red_wine) +
  geom_boxplot(outlier.colour = "transparent",
               color = "#9099A2") +
  coord_flip() +
  geom_jitter(alpha = 0.3,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Alcohol vs Quality",
    y = "Alcohol %",
    x = "Quality score"
    ) + 
  scale_y_continuous(breaks = round(seq(min(red_wine$quality),
                                        max(red_wine$quality), by = 1),1))
finplt1
```

Alcohol compared to quality shows a positive correlation. The values
are dominantly within the 6 and 5 scores. Lowest scoring wines often have
lower alcohol values. While the highest scores tend towards the higher values.
A key observation is category six which shows a range of values across the
alcohol values. This puts into question the strength of this correlation
that higher alcohol equals higher quality of wine.

```{r echo=FALSE, qq2, fig.height = 8, fig.width = 8}
fnl_2 <- function() {
  qqplot((red_wine[red_wine$upper_vs_lower == "upper", ]$alcohol),
  (red_wine[red_wine$upper_vs_lower == "lower", ]$alcohol), 
  pch = 1, 
  xlab = "Alcohol Upper Quality Quantiles", 
  ylab = "Alcohol Lower Quality Quantiles", 
  regress = TRUE,
  main = 'Alcohol % Group QQ plot',
  bty = 'n',  # Frame style
  col = alpha('#192231', 0.3),  # transparency with col
  cex = .4,  # scale pof points
  col.axis = '#9099A2',
  col.lab = '#9099A2',
  col.main = '#192231',
  make.plot = TRUE)
}

fnl_2()
```

Comparing two variables using a quantile quantile plot (Empirical QQ plot) 
can give information about how different the variables are. 
In this case alcohol is split between
upper quality values (>5) and lower quality values (<=5).

The plot and confidence intervals suggest that this the difference between
the two groups of alcohol samples is not significant as the 
confidence bands and values intersects the 1-1 line.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the data-set?

Overall there are few strong correlations between variables in this data-set.
The most promising line of investigation seems to be alcohol and it´s
relationship to quality.

Two other variables (Volatile acidity and sulphates) have a weak relationship
to the feature of interest.

All have been investigated, the challenge relates to the low proportion
of samples in the highest and lowest values compared to the quality
scores of 5 and 6. Often these values show a wider range of values in each
independent variable putting to question if the relationships between
higher and lower quality scores are signal or noise.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Few strong relationships exist, density appears to have associations to
multiple variables and can be investigated further through multivariate
analysis.

A number of plots have outlines, it would be interesting to know if there
are clusters within the data-set or any other types of structures that can
not be observed through bi-variate analysis.

### What was the strongest relationship you found?

Total and free sulfur dioxide have the strongest relationship found,
this would make sense if free sulfur dioxide has
a proportional relationship to the total amount of sulfur
dioxide in a sample.

Density to fixed acidity has the same correlation value in the pair plot
as total and free sulfur dioxide. 

# Multivariate Plots Section

By using the ratio of pairs of correlated independent variables more
dimensions of the data-set can be viewed at the same time. These
paired relationships are based on the bi-variate analysis.

```{r echo=FALSE, Multivariate_Plots_3, fig.height = 20, fig.width = 6, message=FALSE, warning=FALSE}
plt_mv4 <- ggplot(aes(x = (red_wine$density/red_wine$fixed.acidity),
                  y = log10(red_wine$total.sulfur.dioxide /
                         red_wine$free.sulfur.dioxide)
                  ),
       data = red_wine) +
  geom_point(aes(color = factor(red_wine$quality)),
             alpha = 0.5,
             position = position_jitter()
             ) +
    theme_minimal() +
  labs(
    x = "Density g/dm^3/Fixed acidity g/dm^3",
    y = "Total/Free sulfur dioxide mg/dm^3"
    ) +
  labs(color = 'Quality') 
plt_mv4 <- plt_mv4 + scale_colour_brewer(palette = "RdBu")

plt_mv5 <- ggplot(aes(x = (red_wine$density/red_wine$fixed.acidity),
                  y = log10(red_wine$sulphates/red_wine$volatile.acidity)),
       data = red_wine) +
  geom_point(aes(color = factor(red_wine$quality)),
             alpha = 0.5,
             show.legend = FALSE,
             position = position_jitter()
             ) +
    theme_minimal() +
  labs(
    x = "Density g/dm^3/Fixed Acidity g/dm^3",
    y = "Sulphates g/dm^3/Volatile Acidity g/dm^3 to log10"
    )
plt_mv5 <- plt_mv5 + scale_colour_brewer(palette = "RdBu")

plt_mv6 <- ggplot(aes(x = (red_wine$density/red_wine$alcohol),
                  y = log10(red_wine$sulphates/red_wine$volatile.acidity)),
       data = red_wine) +
  geom_point(aes(color = factor(quality)),
             alpha = 0.5,
             show.legend = FALSE,
             position = position_jitter()
             ) +
    theme_minimal() +
  labs(
    x = "Density g/dm^3/Alcohol g/dm^3",
    y = "Sulphates g/dm^3/Volatile Acidity g/dm^3 to log10"
    )
plt_mv6 <- plt_mv6 + scale_colour_brewer(palette = "RdBu")

plt_mv7 <- ggplot(aes(x = (red_wine$pH/red_wine$alcohol),
                  y = log10(red_wine$sulphates/red_wine$chlorides)),
       data = red_wine) +
  geom_point(aes(color = factor(quality)),
             alpha = 0.5,
             show.legend = FALSE,
             position = position_jitter()
             ) +
    theme_minimal() +
  labs(
    x = "pH/Alcohol %",
    y = "Sulphates g/dm^3/Chlorides g/dm^3 to log10"
    )
plt_mv7 <- plt_mv7 + scale_colour_brewer(palette = "RdBu")

multiplot(plt_mv4, plt_mv5, plt_mv6, plt_mv7, cols = 1)
```

Based on bi-variate exploration of variables a number of candidates exist
to explore as ratios. Ratios allow for combining two variables with a
relationship. This can allow for more multivariate exploration, each plot
will combine two pairs of variables to try and see if the quality of wine
stands out in some sort of pattern or relationship.

The upper right plot uses the relationship between total and free sulfur
dioxide and the relationship between density and fixed acidity. Overall
there is weak suggestions that this is aiding in operating quality values.
The are some weak clusters (e.g. value 8) between bottom left and top right.
This is likely reflecting that sulfur dioxide is not having a considerable
impact on the quality value.

The top right plot uses sulphates and volatile acidity transformed to log10
on the y axis with density and alcohol on the x axis. Both pairs of
variables have some correlation and each of these variables has been
previously noted as being correlated to quality. This plot does show a
relationship to quality values with higher quality values being associated
to a low density/alcohol ratio and a positive sulphates/volatile acidity
ratio. This is a noisy relationship with a lot of overlap, especially
from values 5 and 6 which are the most populated.

The lower left plot is similar to the last plot described but swaps
density/alcohol ratio for density/fixed acidity. This shows a similar
relationship as the previous plot but not as strongly along the x axis.

The lower right plot uses a ratio of sulphates and chlorides with a log10
transformation on the y axis with a ratio of pH and alcohol on the x axis.
This also shows a relationship where a high ratio of sulphates and chlorides
along with a lower ratio of pH and alcohol suggest more higher quality wines.

```{r echo=FALSE, Multivariate_Plots_7, fig.height = 6, fig.width = 9}
plt_mv7 <- ggplot(aes(x = (red_wine$density/red_wine$alcohol),
                  y = log10(red_wine$sulphates/red_wine$volatile.acidity)),
       data = red_wine) +
  geom_point(alpha = 0.01,
             position = position_jitter()
             ) +
  theme_tufte() +
  labs(
    title = "Attempting to seperate upper from lower quality wines",
    x = "Density g/dm^3/Alcohol %",
    y = "Sulphates g/dm^3/Volatile Acidity g/dm^3 to log10"
    )
fnl_plt_4 <- plt_mv7 + geom_point(aes(size = quality), 
                                  pch = 21, 
                                  show.legend = T,
                                  alpha = 0.5) + 
  aes(fill = factor(upper_vs_lower)) +
  labs(fill = 'Upper vs Lower Quality') 
fnl_plt_4
```

Based on multivariate exploration two pairs of ratios are selected and
plotted with upper and lower quality categories as colour and quality as
the size in this bubble plot. The size of the shape is subtle due to the
dominance of values 5 and 6. It helps highlight the lowest scoring values.

This plot shows there is relationship within the data that can begin to
separate some of the lower and upper quality wines. It also shows there
is significant overlap.

```{r echo=FALSE, Multivariate_Plots_8, fig.height = 6, fig.width = 9}
plt_mv8 <- ggplot(aes(x = (red_wine$alcohol),
                  y = log10(red_wine$sulphates/red_wine$volatile.acidity)),
       data = red_wine) +
  geom_point(alpha = 0.1,
             position = position_jitter()
             ) +
  theme_minimal() +
  labs(
    title = "Comparing ratios of key variables",
    x = "Alcohol %",
    y = "Sulphates g/dm^3/Volatile Acidity g/dm^3 to log10"
    )
plt_mv8 <- plt_mv8  + facet_wrap(~ factor(red_wine$quality))
plt_mv8 <- plt_mv8 + geom_point(
                     size = 2, 
                     pch = 21,
                     show.legend = T) + 
  aes(fill = factor(quality), alpha = 0.2) +
  labs(fill = 'Quality') 
plt_mv8
```

The last plot uses a chart separated by each factor of quality. The
axis are based on the same pair of ratios as previously used to highlight
differences between upper and lower quality wines against alcohol.

The main purpose of this plot is to again show the challenges of the number
of samples outside of values 5 and 6. Having a more even sample of values
in the higher or lower values would increase the confidence that relationship
previously described are reasonable and can be pursued further.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

To assess the dependent variable quality
the most productive investigation was using ratio of
correlated variables.

The chosen pair of density/alcohol and sulphates/volatile acidity (using a
log10 transform) proved most effective at separating upper and lower
quality wines. Neither of these pairs have the strongest correlation
but both pairs showed a good distribution of values which may be why
they work well as ratios to describe variations in the data-set.

### Were there any interesting or surprising interactions between features?

Despite selecting four variables to highlight the relationship different
combinations of variables each gave different insights into the data-set.

It was surprising to see the relationship using upper and lower quality
considering that bi-variate and uni variate analysis has so far only
shown weak relationships.

Principal component analysis could be used to expand this investigation.

### Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

No models have been created due to ambiguity between the dependent and
independent variables. Machine learning could create a model
based on the observations.

Trying to get a better sample of across quality values should be prioritized
before trying sophisticated modelling in this case.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One, fig.height = 8, fig.width = 8}
finplt1
```

### Description One
The dependent variable within this data-set is quality measured between 0-10.
The majority of samples lie between scores of 5 and 6. 2 and 3 is very
poorly sampled and 2 to 0, 9 & 10 do not occur in the data-set. This makes
it challenging to identify robust relationships between independent variables.
This plot highlights that there is no information outside of the range 3-8.

Alcohol is one of the variables which have an apparent relationship with
quality. At first look there appears to be a trend suggesting lower alcohol,
lower quality and vice versa.

### Plot Two
```{r echo=FALSE, Plot_Two, fig.height = 8, fig.width = 8}
fnl_2()
```

### Description Two
Comparing two variables using a quantile quantile plot (Empirical QQ plot) 
can give information about if the difference is significant. 
In this case alcohol is split between
upper quality values (>5) and lower quality values (<=5). This is chosen
as alcohol has the strongest correlation to quality (0.48) and the observation
made when comparing frequency polygons between these two groups.

The steps in the plot will relate to rounding of measurement. The majority of
the values between 10% and 12.5% are separated with higher values in
the upper quality group. The 95% confidence bands highlight that this is not
a significant observation as the two tails intersect the 1-1 line.

This plot is chosen to highlight that the independent variable with the
strongest correlation does not show a significant relationship between
upper and lower quality groups.

### Plot Three
```{r echo=FALSE, Plot_Three, fig.height = 6, fig.width = 10}
fnl_plt_4
```

### Description Three
Based on multivariate exploration of many combinations of pairs of ratios
two are selected to highlight trends in the data. 
Plotted with upper and lower quality categories as colour and quality as
the size in this bubble plot. The size of the shape is subtle due to the
dominance of values 5 and 6. It helps highlight the lowest scoring values.

The two ratios used are density/alcohol along the x axis and 
sulphates/volatile acidity with a log10 transform. 
There are weak relationship between each of these pairs and all variables
have been seen to have weak to moderate impact on the dependent quality
variable.

This plot shows there is relationship within the data that can begin to
separate some of the lower and upper quality wines. Machine learning
approaches could fit a model to separate these groups of values.
It also shows there is significant overlap, which will hamper the performance
of any model created. Expectations should not be too high about having any
high quality predictive model from this data-set.
------

# Reflection

By systematically working through all of the variables in a uni-variate and
bi-variate analysis this allowed for the multivariate analysis to be steered
in a way that has shown a relationship to the dependent variable.

This was much more effective than just playing with advanced multivariate
graphs to plot many variables which is often a temptation.

The struggle with this data-set was not having any domain knowledge about the
topic and then the lack of correlated variables. Perhaps there are ways to
better identify a relationship between quality and other variables with
more knowledge about the measurement types, however I feel this analysis
has explored most of the potential combinations of variables.

The other key issue was the low sample rate of quality outside of 5 and 6.
This data-set is a good case where gathering a broader sample of observations
should be prioritized before committing time to other methodologies
like machine learning.

The main proposal for further work would be to focus on getting a more
even representation of data for all labels. There is a similar dataset
for white wine which could be compared to see if there are any
similar relationships between this trend. This is still a limited sample
as we are looking at only one region of wine.

Principal component analysis could be used to reduce the dimensionality of the
dataset and explore if this relates to qualiy.

The quality measure itself could also be verified as an experts opinion on wine
may be subjective. Perhaps collecting each individual reviewers score for
each sample will shed further insight.

After exploring further an experiment would be to make a
hypothesis and what
controls wine quality, and then test this on new larger datasets. 
If this includes
attributes that can be controlled for during production of wine this may aid
in wine producers producing better quality wine.

For example are there a typical 
range of alcohol values that are preferred by wine
experts and customers.

## Dataset from:
P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
Modeling wine preferences by data mining from physicochemical properties.
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.
Available at: 
[@Elsevier](http://dx.doi.org/10.1016/j.dss.2009.05.016)
[Pre-press (pdf)](http://www3.dsi.uminho.pt/pcortez/winequality09.pdf)
[bib](http://www3.dsi.uminho.pt/pcortez/dss09.bib)